Restrictive Clearance Comparison

### 2025-01-10 data-raw/RestrictiveClearance_comparison.Rmd
Added diagrams for chemical list overlap.

```{r library_loading}
today <- format(Sys.Date(), "%d%b%Y")
devtools::load_all()
library(tidyverse)
library(ctxR)
library(ggvenn)
source("data-raw/RC_comparison_bkgd.R") # httk loaded here
# Only use plasma 
this_pk <- pk(cvt) +
  settings_optimx(method = "bobyqa") +
  facet_data(vars(Chemical, Species)) +
  settings_preprocess(
    keep_data_original = TRUE,
    suppress.messages = FALSE
  ) +
  scale_conc(dose_norm = TRUE, log10_trans = TRUE) +
  stat_error_model(error_group = vars(Chemical, Species))
this_pk <- do_fit(this_pk, n_cores = 12)
```

@@

member_list <- list(
  httk_gas_pbtk = unique(get_cheminfo(info = "DTXSID",
                                species = "rat",
                                model = "gas_pbtk",
                                median.only = TRUE,
                                physchem.exclude = TRUE,
                                suppress.messages = TRUE)) %>% 
    suppressWarnings(),
  cvt = cvt %>% filter(species %in% c("rat")) %>% pull(analyte_dtxsid)
)

ggsave(filename = paste0(Sys.getenv("FIG_DIR"),
                         "RestrictiveClearance_Analysis/",
                         today, "httk_vs_CvT_VennDiagram_ratOnly.png"),
       bg = "white",
       height = 4, width = 6
)

new_chems <- clipr::read_clip()
length(unique(new_chems %in% setdiff(member_list$cvt, member_list$httk_gas_pbtk)))
length(unique(new_chems %in% member_list$cvt))

intersect(setdiff(member_list$cvt, member_list$httk_gas_pbtk), new_chems)
```

```{r combine_tkstats}
# Start Here if you ran the background job

my_tkstats <- eval_tkstats(my_pk, finite_only = FALSE) %>%
  filter(Media %in% "plasma",
         Chemical %in% this_chem_v,
         Species %in% this_species_v) %>%
  semi_join(get_winning_model(my_pk)) %>%
  dplyr::select(Chemical, Species, CLtot.tkstats, CLtot.nca) %>%
  distinct() %>%
  filter(
    !is.na(CLtot.tkstats), !is.na(CLtot.nca)
  ) %>%
  group_by(Chemical, Species, CLtot.tkstats) %>%
  summarize(
    CLtot.nca_mean = mean(CLtot.nca),
    CLtot.nca_sd = sd(CLtot.nca)
  )

# Combine these two
combined_clearance_df <- this_httk_clearance_df %>%
  inner_join(my_tkstats) %>%
  distinct()

@@

httk_preds_rat <- get_httk_preds(rat_pars_rat, my_pk, species = "rat")

@@

```{r rat_data}
rat_df <- httk_preds_rat %>%
  left_join(predict(my_pk)) %>%
  semi_join(get_winning_model(my_pk))

rat_df <- ggplot(mapping = aes(
    x = Conc_est,
    y = httk_Preds,
    color = interaction(Route, Media)
  )) +
  geom_point() +
  geom_line(mapping = aes(
    group = interaction(Chemical, Media, Route, Reference)
  )
  ) +
  geom_abline(slope = 1, intercept = 0, color = "black") +
  coord_equal(ylim = c(0, 200), xlim = c(0, 200))

@@

saveRDS(new_pk, "data-raw/ratCVT_wHTTK_reformulated.rds")
```

@@

```{r begin_analysis}
new_pk <- readRDS("data-raw/ratCVT_wHTTK_reformulated.rds")

coef(new_pk) %>% View()
eval_tkstats(new_pk, finite_only = FALSE) %>% View()
get_winning_model(new_pk) %>%
  count(model)

@@

rel_ll %>%
  rowwise() %>% 
  mutate(across(contains("_1comp"),
                \(x) {
                  ifelse(
                    x < model_flat | x < 1,
                    NA_real_,
                    x
                  )
                })) %>% 
  count(is.na(model_1comp_fup))
  # drop_na(model_1comp_fup) %>%
  # filter(Chemical %in% clear_chem_levels) %>% View()
  glimpse()

new_pk$fit %>%
  filter(param_name %in% "Fup",
         model %in% "model_1comp_fup") %>%
  ggplot(aes(x = estimate, y = start)) +
  geom_point(size = 1.25) +
  geom_point(data = new_pk$fit %>%
               filter(param_name %in% "Fup",
                      model %in% "model_1comp_fup",
                      Chemical %in% combined_clearance_df$Chemical),
             size = 1.5, color = "green4") +
  theme_bw() +
  coord_equal(xlim = c(0,1), ylim = c(0,1),
              expand = TRUE) +
  coord_flip()

@@

For this I will just use the PK object I have saved `my_pk` and take the
_prefit_ parameter estimations and first compare that with eventual optimized `kelim`.

```{r kelim-questions}
# Original data, filtered for common_chems
# par_DF
my_kelim_prefit <- my_pk$prefit$par_DF %>%
  filter(
    param_name %in% "kelim",
  ) %>%
  distinct(Chemical, Species, param_name, lower_bound, upper_bound, start)

my_kelim_fit <- get_fit(my_pk) %>%
  semi_join(get_winning_model(my_pk)) %>%
  ungroup() %>%
  filter(param_name == "kelim") %>%
  distinct(model, Chemical, Species, param_name, value)

my_kelim_df <- inner_join(my_kelim_fit, my_kelim_prefit)

@@

full_df <- inner_join(my_kelim_df, httk_df) %>%
  relocate(fit_kelim, .after = last_col())

@@

my_data <- get_data(new_pk) %>%
  distinct(Chemical, Species, Reference, Route, Media, Dose, Time, Conc) %>%
  right_join(new_resid)

@@

```{r oxybenzene-test}
old_AIC <- AIC(my_pk) %>%
  filter(
    Chemical %in% "DTXSID3022405",
    Species %in% "mouse"
  )

@@

old_rsquared <- rmse(my_pk) %>%
  filter(
    Chemical %in% "DTXSID3022405",
    Species %in% "mouse"
  )

@@

my_pl_1comp <- plot(new_pk,
                    n_interp = 11, use_scale_conc = FALSE,
                    model = c(
                      "model_1comp",
                      "model_1comp_rest",
                      "model_1comp_nonrest"
                    )
)

my_pl_2comp <- plot(new_pk,
                    n_interp = 11, use_scale_conc = FALSE,
                    model = c(
                      "model_2comp",
                      "model_2comp_rest",
                      "model_2comp_nonrest"
                    )
)

@@

new_cl_tkstats <- get_tkstats(new_pk)
my_tkstats <- my_pk %>%
  get_winning_model() %>%
  left_join(coef(my_pk)) %>%
  left_join(rsq(my_pk)) %>%
  distinct()
my_tkstats %>%
  select(-coefs_vector) %>%
  clipr::write_clip()
# Send the R2 too.

>>>>

```{r library_loading}
today <- format(Sys.Date(), "%d%b%Y")
devtools::load_all()
library(tidyverse)
library(ctxR)
library(ggvenn)
library(ggalluvial)
library(reticulate)
source("data-raw/RC_comparison_bkgd.R") # httk loaded here
```

```{r quick-fit}
# Only use plasma 
this_pk <- pk(cvt) +
  settings_optimx(method = "bobyqa") +
  facet_data(vars(Chemical, Species)) +
  settings_preprocess(
    keep_data_original = TRUE,
    suppress.messages = FALSE
  ) +
  scale_conc(dose_norm = TRUE, log10_trans = TRUE) +
  stat_error_model(error_group = vars(Chemical, Species))
this_pk <- do_fit(this_pk, n_cores = 12)
```

@@

# new_chems <- clipr::read_clip()
# length(unique(new_chems %in% setdiff(member_list$cvt, member_list$httk_gas_pbtk)))
# length(unique(new_chems %in% member_list$cvt))

# intersect(setdiff(member_list$cvt, member_list$httk_gas_pbtk), new_chems)

target_chems <- intersect(member_list$cvt, member_list$httk_gas_pbtk)
```

Here I quickly check a few additional chemicals that will be added to _httk_.

```{r new_httk_chems}
new_chems_clint <- readxl::read_xlsx(path = paste0(Sys.getenv("OD_DIR"),
                                             "162Clint_Com_ChemID_DTXSID.xlsx"))
new_smiles <- readxl::read_xlsx(path = paste0(Sys.getenv("OD_DIR"),
                                             "127NewSMILES_Caco2_Cyprotex_7.23.24.xlsx"))
# New Clint
intersect(
  member_list$cvt[which(!(member_list$cvt %in% member_list$httk_gas_pbtk))], new_chems_clint$DTXSID
  )

# New SMILES
intersect(
  member_list$cvt[which(!(member_list$cvt %in% member_list$httk_gas_pbtk))], new_smiles$DTXSID
  )
```

## Lipinski's Rules

Here I will look at the whether chemicals are "drug-like" by evaluating if:

1. Molecular weight less than 500 Da.  
2. logP (octanol-water partitioning coefficient) no greater than 5.  
3. Less than 5 hydrogen bond donors and less than 10 acceptors.

```{r lipinksi_test_data}
target_chem_info <- get_chemical_details_batch(DTXSID = target_chems,
                                               rate_limit = 1L)
target_chem_properties <- get_chem_info_batch(DTXSID = target_chems, 
                                              rate_limit = 1L) %>%
  filter(name == "LogKow: Octanol-Water") %>% distinct()

target_chem_properties %>%
  ggplot(aes(x = source, y = value)) +
  geom_point() +
  geom_line(aes(group = dtxsid), color = "grey10", alpha = 0.1) +
  geom_hline(yintercept = 5, linetype = "longdash") +
  facet_grid(cols = vars(propType), scales = "free_x") +
  theme_bw() +
  theme(axis.text.x = element_text(hjust = 1, angle = 45))

target_chem_properties %>%
  count(source)

target_chem_info %>%
  ggplot(aes(x = monoisotopicMass)) +
  geom_histogram(binwidth = 20, fill = "grey5") +
  geom_vline(xintercept = 500, linetype = "longdash") +
  theme_bw()

target_chem_info %>%
  count(monoisotopicMass <= 500)

```{r reticulate-prep}
smiles_list <- target_chem_info %>% pull(smiles) %>% unique()
mykit <- import("rdkit")
```

```{python get_H_bonds}
import rdkit
from rdkit import Chem
from rdkit.Chem import Descriptors

py_smiles_list = r.smiles_list
mol_from_smiles = [Chem.MolFromSmiles(x) for x in py_smiles_list]

py_h_donors = [Chem.Descriptors.NOCount(x) for x in mol_from_smiles]
py_h_acceptors = [Chem.Descriptors.NHOHCount(x) for x in mol_from_smiles]
```

```{r lipinksi_alluvial}
target_chem_summary <- target_chem_info %>%
  distinct(dtxsid, monoisotopicMass, smiles) %>%
  right_join(data.frame(smiles = smiles_list)) %>%
  bind_cols(data.frame(H_donors = py_to_r(py$py_h_donors),
                       H_acceptors = py_to_r(py$py_h_acceptors))
            ) %>%
  left_join(
    target_chem_properties %>%
      filter(source == "ACD/Labs Consensus") %>%
      distinct(dtxsid, source, value)
  ) %>%
  rename(LogKow = value)

target_chem_lipinski <- target_chem_summary %>%
  select(-c(smiles, source)) %>%
  group_by(dtxsid) %>%
  mutate("Mass <= 500 Da" = (monoisotopicMass <= 500),
          "H-donors < 5" = (H_donors < 5),
          "H-acceptors < 10" = (H_acceptors < 10),
          "logP < 5" = (LogKow < 5),
          .keep = "none") %>%
  ungroup() %>%
  mutate(Lipinski = rowSums(dplyr::select(., where(is.logical))) >= 3)

target_chem_lipinski %>%
  group_by(Lipinski) %>%
  count(`Mass <= 500 Da`, `H-donors < 5`, `H-acceptors < 10`, `logP < 5`,
        name = "Count") %>%
  ggplot(aes(
    axis3 = `Mass <= 500 Da`,
    axis4 = `logP < 5`,
    axis2 = `H-donors < 5`,
    axis1 = `H-acceptors < 10`,
    y = Count
  )) +
  geom_alluvium(aes(fill = Lipinski)) +
  geom_stratum() +
  scale_x_discrete(limits = c( 
    "H-acceptors < 10",
    "H-donors < 5",
    "<= 500 Da",
    "logP < 5"
  ),
  position = "top",
  expand = c(.2, .05)) +
  geom_text(stat = "stratum", aes(label = after_stat(stratum))) +
  labs(title = "Overwhelmingly, this subset of chemicals passes Lipinksi's Rules of 5") +
  theme_minimal() +
  theme(axis.text.x = element_text(face = "bold", color = "black"),
        panel.grid = element_blank())

Overwhelmingly, we see that chemicals (by Lipinski's rules of 5) are "drug-like".
While this may not matter effectively for clearance, it may be nice to look 
at whether chemicals deemed not "drug-like" (11 total) are more likely to be
restrictively or non-restrictively cleared.

```{r combine_tkstats}
# Start Here if you ran the background job
this_winmodel <- get_winning_model(this_pk)
this_tkstats <- inner_join(this_winmodel %>% filter(model != "model_flat"),
                          eval_tkstats(this_pk, finite_only = TRUE)) %>%
  filter(Media %in% "plasma",
         Chemical %in% this_chem_v,
         Species %in% this_species_v,
         !is.na(CLtot.tkstats)) %>%
  semi_join(get_winning_model(this_pk)) %>%
  dplyr::select(Chemical, Species, CLtot.tkstats, CLtot.nca) %>%
  distinct() %>%
  group_by(Chemical, Species, CLtot.tkstats) %>%
  summarize(
    CLtot.nca_mean = mean(CLtot.nca, na.rm = TRUE),
    CLtot.nca_sd = sd(CLtot.nca, na.rm = TRUE)
  )
message(length(unique(this_tkstats$Chemical)),
        " unique chemicals with finite clearance.")

# Combine these two
combined_clearance_df <- this_httk_clearance_df %>%
  inner_join(this_tkstats) %>%
  distinct()

@@

httk_preds_rat <- get_httk_preds(rat_pars_rat, this_pk, species = "rat")

@@

```{r rat_data}
rat_df <- httk_preds_rat %>%
  left_join(predict(this_pk)) %>%
  semi_join(this_winmodel)

rat_df %>% ggplot(mapping = aes(
    x = Conc_est,
    y = httk_Preds,
    color = interaction(Route)
  )) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, color = "black") +
  scale_x_log10() + scale_y_log10() +
  facet_wrap(~Route)

@@

# saveRDS(new_pk, "data-raw/ratCVT_wHTTK_reformulated.rds")
```

@@

```{r begin_analysis}
# new_pk <- readRDS("data-raw/ratCVT_wHTTK_reformulated.rds")
coef(new_pk) %>% View()
new_winmodel <- get_winning_model(new_pk)
eval_tkstats(new_pk, finite_only = TRUE) %>% 
  semi_join(new_winmodel) %>%
  View()

@@

rel_ll %>%
  rowwise() %>% 
  mutate(across(contains("_1comp"),
                \(x) {
                  ifelse(
                    x < model_flat | x < 1,
                    NA_real_,
                    x
                  )
                })) %>% 
  count(is.na(model_1comp_fup))
  # drop_na(model_1comp_fup) %>%
  # filter(Chemical %in% clear_chem_levels) %>% View()

new_pk$fit %>%
  filter(param_name %in% "Fup",
         model %in% "model_1comp_fup") %>%
  ggplot(aes(x = estimate, y = start)) +
  geom_point(size = 1.25) +
  theme_bw() +
  coord_equal(xlim = c(0,1), ylim = c(0,1),
              expand = TRUE) +
  coord_flip()

@@

For this I will just use the PK object I have saved `this_pk` and take the
_prefit_ parameter estimations and first compare that with eventual optimized `kelim`.

```{r kelim-questions}
# Original data, filtered for common_chems
# par_DF
this_kelim_prefit <- this_pk$prefit$par_DF %>%
  filter(
    param_name %in% "kelim",
  ) %>%
  distinct(Chemical, Species, param_name, lower_bound, upper_bound, start)

this_kelim_fit <- get_fit(this_pk) %>%
  semi_join(get_winning_model(this_pk)) %>%
  ungroup() %>%
  filter(param_name == "kelim") %>%
  distinct(model, Chemical, Species, param_name, value)

this_kelim_df <- inner_join(this_kelim_fit, this_kelim_prefit)

@@

full_df <- inner_join(this_kelim_df, httk_df) %>%
  relocate(fit_kelim, .after = last_col())

@@

this_data <- get_data(new_pk) %>%
  distinct(Chemical, Species, Reference, Route, Media, Dose, Time, Conc) %>%
  right_join(new_resid)

@@

```{r oxybenzene-test}
old_AIC <- AIC(this_pk) %>%
  filter(
    Chemical %in% "DTXSID3022405",
    Species %in% "mouse"
  )

@@

old_rsquared <- rmse(this_pk) %>%
  filter(
    Chemical %in% "DTXSID3022405",
    Species %in% "mouse"
  )

@@

this_pl_1comp <- plot(new_pk,
                    n_interp = 11, use_scale_conc = FALSE,
                    model = c(
                      "model_1comp",
                      "model_1comp_rest",
                      "model_1comp_nonrest"
                    )
)

this_pl_2comp <- plot(new_pk,
                    n_interp = 11, use_scale_conc = FALSE,
                    model = c(
                      "model_2comp",
                      "model_2comp_rest",
                      "model_2comp_nonrest"
                    )
)

@@

new_cl_tkstats <- get_tkstats(new_pk)
this_tkstats <- this_pk %>%
  get_winning_model() %>%
  left_join(coef(this_pk)) %>%
  left_join(rsq(this_pk)) %>%
  distinct()
this_tkstats %>%
  select(-coefs_vector) %>%
  clipr::write_clip()
# Send the R2 too.
